---
title: "Confounding of Selection and Influence"
author: "Cecilia Cavero-Sanchez"
date: "`r Sys.Date()`"
output: pdf_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

-   There are important challenges in conducting causal inference on contagion effects in observational data.

-   [@fyfe2024] show how we can use the "split-haves" test, robust to confounding, and apply it to studies of contagion effects.

-   In this tutorial I will go over the method and several replication examples.

## The split halves test

-   Observational data are subject to confounding when identifying contagion/influence effects because of the co-existence of homophily and influence.

-   The SH test isolates the impact of contagion by assuming the pre-existence of a network in the data without conditioning on it.

    1.  Test data and adjust it for non-stationarity.
    2.  Randomly split observational time-series cross-section data into two halves based on node (country in country-year data).
    3.  Calculate mean values for each half for every time period.
    4.  Run regression setting time $t$ means as the dependent variable and $t-1$ means of each half as independent variables.
    5.  Perform steps $1-3$ $N$ times to recover a mean and $p-$value that indicate whether contagion is present or not.
    6.  Contagion signal is the average value of the estimated relationship between the mean value of the first half at time $t$ with the mean value of the second half at time $t-1$, conditional on the mean value of the first half at $t-1$. In a way, it is the relationship between both halves at different time pints.
    7.  The $p-$value is calculated as the minimum of two proportions, the proportion of times the contagion signal is $>0$ and the proportion of times when it is $>0$. We obtain the $p-$value by multiplying the minimum proportion by $2$ for a two-tailed test of whether there is contagion in the data.
    8.  The estimate of general contagion tells us the average effect of a one-unit increase in the outcome value of any other node in the following year.

## Applying the split-halves test

-   I will use three of the replication examples in [@fyfe2024] to illustrate the use of the split-halves test and its impact on results of previous studies.

### Confirmation of Contagion: Conflict Onset

-   The first replication is of [@buhaug2008], who find that there is a neighborhood effect of armed conflict.

-   DV is binary indicator of conflict onset and a three-level ordinal variable indicating the type of conflict.

-   First, here is the replication of the original study.

```{r buhaug and gleditsch, warning = F, message = F}
## Buhaug and Gleditsch
rm(list = ls())
# Libraries
library(ggplot2)
library(dplyr)
library(broom)
library(nnet)
library(dplyr)
library(haven)

# Original data
rep <- read_dta("hb_ksg_replication.dta")

# Model 1
model1 <- glm(allons3 ~ neighall + neighpol + I(neighpol^2) + 
                neighlgdp + peaceall, 
              data = rep, family = binomial(link = "logit"))

# Model 2
model2 <- glm(allons3 ~ ncivwar + neighpol + I(neighpol^2) + 
                neighlgdp  + peaceall, 
              data = rep, family = binomial(link = "logit"))

# Model 3
model3 <- glm(allons3 ~ neighall + neighpol + I(neighpol^2) + 
                neighlgdp + polity2l + I(polity2l^2) + 
              lgdp96l + lnpop + peaceall, 
              data = rep, family = binomial(link = "logit"))

# Model 4
model4 <- glm(allons3 ~ ncivwar + neighpol + I(neighpol^2) + 
                neighlgdp + polity2l + I(polity2l^2) + 
              lgdp96l + lnpop + peaceall, 
              data = rep, family = binomial(link = "logit"))

# Table 2, conflict neighbors only
# Model 5
model5 <- glm(allons3 ~ lnblength + lndist + ethlink2 + 
                lneighbref + pop_nc + nterr + lbd_cum + 
                polity2l + I(polity2l^2) + lgdp96l + lnpop +
                peaceall, 
              data = subset(rep, ncivwar == 1), 
              family = binomial(link = "logit"))

# Model 6
model6 <- glm(allons3 ~ lnblength + confbord + ethlink2 + 
                lneighbref + pop_nc + nterr + lbd_cum + 
              polity2l + I(polity2l^2) + lgdp96l + lnpop + 
                peaceall, 
              data = subset(rep, ncivwar == 1), 
              family = binomial(link = "logit"))

# Model 7, multinomial logit of terr and gov conflicts
model7 <- multinom(mons3 ~ ethlink2 + lneighbref + nterr +
                    polity2l + I(polity2l^2) + 
                     lgdp96l + lnpop + peaceall, 
                   data = subset(rep, ncivwar == 1))
# Summary
library(texreg)
screenreg(list(model1, model2, model3, model4))
screenreg(list(model5, model6, model7))
```

-   The results from models $1-4$ (full sample analysis) show that the presence of neighboring conflicts (neighall) and the number of neighboring civil wars (ncivwar) significantly increases the likelihood of conflict onset.

-   The results from models $5-7$ show that ethnic links to neighboring conflicts (ethlink2) and the number of territorial conflicts in neighboring countries (nterr) significantly increases the risk of conflict spillover.

-   Below, by applying the split halves test, we see that this is an example where the SH test confirms the author's main results of contagion.

```{r split halves buhaug gleditsch, warning = F, message = F}
## Replication
# Libraries
library(maditr)
library(ggplot2)
library(tidyr)
library(ggplot2)
library(haven)
library(ContagionTest) # can download from GitHub
#devtools::install_github("rebekahfyfe/ContagionTest")

# Data
d <- read.table("conflict.tab", header = T)

# Removing duplicate rows
d <- d[-4439,]
d <- d[-5014,]

# Selecting necessary columns (country, date, DV)
d1 <- d %>%
  select("abbrev", "year", "allons3") %>%
  pivot_wider(names_from = year, values_from = allons3)
d1 <- as.data.frame(d1)

# Changing null values to NAs (treating as missing data)
d1[d1 == 'NULL'] <- NA

# Formatting for split-halves test
d1 <- STFormat(d1)

# Running split-halves contagion test
simmodels <- lag_pc_test(d1, 1000, 3, T, 0.1, 
                         lagWin = 1, missingData = T)
# Summary of models
simmodels <- as.data.frame(simmodels)
names(simmodels) <- c("intercept","t-1coef","counterpart")

# Calculate contagion signal
xmean <- mean(simmodels$counterpart) ## input this in the plot
xmean <- round(xmean, digits = 4)

# P-value of the signal (proportion of results < 0)
pval <- sum(simmodels$counterpart < 0) / 1000  ## pvalue
pval <- round(pval, digits = 3)

# Density graph of results
density_graph(simmodels, 1000, xmean, 0.5, xmean, 0.25, 
              title = "Global Conflict")
```

-   Using $1,000$ random splits, the SH returns a positive contagion signal and a $p-$value $<0.01$.

-   The expected prevalence of civil conflict onset in one country increases by approximately $0.02$ for every $0.1$ increase in lagged civil war prevalence among the other countries.

-   The results support the findings of Buhaug and Gleditsch.

### Challenging Non-Contagion: Pro-Democracy Protests

-   The authors find that, contrary to some strands of literature, pro-democracy protests do not diffuse to other countries [@brancati2018].

-   DV is protest onset.

-   This is an example where the SH test challenges the author's results of no contagion.

```{r pro-democracy protests, warning = F, message = F}
## Brancati and Lucardi
# Violence in the Netherlands data, from Braun 2011
d <- as.data.frame(read.delim(unzip("violneth.zip")))

# Selecting only the necessary columns
d <- d[, c(3, 5, 21:55)]

# Creating a singe variable for dates
T1 <- c(rep(1, 30), rep(0, 1065))
d$TT1 <- rep(T1, 474)
date <- seq(as.Date("2001-01-01"), 
            as.Date("2003-12-31"), by = "days")
d$date <- rep(date, 474)

# Changing to wide format
d <- d %>% select(number, date, countinc) %>%
  pivot_wider(names_from = date, values_from = countinc)

# Formatting to be used with congation test
d <- STFormat(d)

# Running parallel contagion test
simmodNVio <- lag_pc_test(d, 1000, 1, T, 0.05)
summary(simmodNVio)

# Creating a dataframe with results
simmodNVio <- as.data.frame(simmodNVio)
names(simmodNVio) <- c("intercept","t-1coef","counterpart")

# Calculating mean (contagion signal)
mean <- mean(simmodNVio$counterpart) ## input this in the plot below
(mean <- round(mean, digits = 10))
# Significance of the signal, proportion of means less than 0
pval <- sum(simmodNVio$counterpart < 0) / 1000  ## pvalue
(pval <- round(pval, digits = 3))

# Density graph of results
density_graph(simmodNVio, 1000, mean, 1, mean, 0, 
              title = "Ethnic Violence in the Netherlands")
```

-   While Brancati and Lucardi find no contagion effect, the SH test shows that there is indeed statistical evidence of contagion.

### Challenging Contagion: Civilian Targeting

-   The authors find that there is a spillover effect that results in the spreading of violence against civilians by armed actors [@lis2021].

-   DV is the Civilian Targeting Index, which ranges from $0$ to $1000$.

-   This is an example where the SH test challenges the author's results of contagion.

```{r civilian targeting, warning = F, message = F}
## Replication data
d <- read_dta("civiliantargeting.dta")

# Selecting necessary columns (country, date, DV)
d <- d %>%
  select(actor_id, year, cti)

# Changing to wide format
d <- dcast(d, actor_id ~ year, value.var = "cti")

# Formatting to be used with split-halves contagion test
d <- STFormat(d)

# running contagion test
lslres <- lag_pc_test(d, 1000, 1, T, 0.05, 1, F)
summary(lslres)

# Creating a data frame with results
lslres <- as.data.frame(lslres)
names(lslres) <- c("intercept","t-1coef","counterpart")

# Calculating mean (contagion signal)
lslresmean <- mean(lslres$counterpart) ## input this in the plot below
(lslresmean <- round(lslresmean, digits = 10))
# Significance of the signal, proportion of means less than 0
lslrespval <- sum(lslres$counterpart < 0) / 1000  ## pvalue
(lslrespval <- round(lslrespval, digits = 3))

# Density graph of results
density_graph(lslres, 1000, lslresmean, 0.7, lslresmean, 0.4, 
              title = "Civilian Targeting")
```

-   While Lis, Spagat and Lee find a contagion effect, the SH test shows that there is no statistical evidence of contagion.

\newpage

## References
